---
title: "Creating and Evaluating Seasonal Influenza Thresholds in Philadelphia"
subtitle: "BMIN5030 Final Project"
author: "Hope Schreiber"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers. Feel free to change the theme and other display settings, although this is not required.

## Overview {#sec-overview}

The goal of this project is to explore two popular methods for determining intensity thresholds for seasonal influenza in Pennsylvania. The data utilized will be from the CDC's FluVIew Interactive database \[1\]. This is an online tool that provides publicly available, weekly influenza-like illness (ILI) data at the state level. The overall purpose of this project is to apply the methods to Penn's own community, with the help of Sidney P and Kendra T (Epidemiologist and Data Analyst on the Public Health team at Penn Wellness). They have provided guidance on data organization and visualization, as well as dissemination of results.

[BMIN5030 Final Project GitHub Repository](https://github.com/HopeSch01/BMIN5030_Final_Project_Schreiber)

## Introduction {#sec-introduction}

To capture influenza (flu) season severity both actively and retrospectively, the Center for Disease Control and Prevention (CDC) uses intensity thresholds \[2\]. An intensity threshold is established from data gathered during previous flu seasons, and is then used to assess how severe the current flu season is in comparison \[2\]. They help researchers decide how severe flu activity is during peak flu season, and how severe the season was as a whole once it is finished. This information helps to guide public health action such as targeted messaging and prevention measures.

There is not a "gold standard" for calculating intensity thresholds, and multiple methods exist. In 2012, the World Health Organization (WHO) established a new method for defining flu flu thresholds \[3\]. It utilizes means and standard deviation from historical flu data, and the WHO released a standard guide for determining seasonal and alert thresholds \[4\]. The moving epidemic model (MEM) is another method for calculating influenza intensity thresholds. It also uses historical data, but involves more statistics and relies heavily on consecutive seasons of data \[5\]. An R package has been created to guide MEM use \[6\].

The CDC collaborates with state health departments and other clinical laboratories to maintain the U.S. influenza surveillance system \[1\]. Information on outpatient visits is reported by partners through the Influenza-like Illness Surveillance Network (ILINet) \[1\]. More than 110 million patient visits were reported during the 2022-23 season \[1\]. This is where the Public Health Team at Penn Wellness reports weekly ILI cases among students. With robust historical data available, seasonal flu thresholds can be established in Pennsylvannia to better understand current flu activity. Additionally, this provides an opportunity to compare two popular methods in establishing thresholds. This is a project that is rooted in the core epidemiological task of surveillance, but also includes statistical modeling skills and relevance to virology. Information from this project can help inform seasonal flu thresholds on UPenn's own campus. The Public Health Team at Penn Wellness collects weekly flu data, and maintains historical flu data spanning back to 2009. There is an opportunity for this project to have a real impact on our own campus community.

## Methods {#sec-methods}

Describe the data used and general methodological approach used to address the problem described in the @sec-introduction. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.

The data utilized for this project is from the U.S. influenza surveillance system in collaboration with the CDC \[1\]. All data used in this project is at the state level, though FluView has data available at the regional and national level.

In order to create seasonal influenza thresholds, data specifically from ILINet was utilized. This includes information on outpatients visits to health care providers for respiratory illness. This setting most closely aligns to that of Penn students, which is why it was chosen for this project.

An ILI case is defined in this system as "a fever (temperature of 100°F \[37.8°C\] or greater) and a cough and/or sore throat. ILINet will capture any visit due to any respiratory pathogen that presents with this case definition.

FluView includes other important surveillance data including age group distribution of positive influenza tests, hospitalizations, and mortality. These will also be explored to obtain a complete and accurate picture of ILI activity prior to establishing state level seasonal activity thresholds.

#### Intro/background Exploration

-   line graph of ILI cases by CDC week over the years

    -   ILINet data

-   map showing change in ILI activity?

    -   would need to download data for all states from FluView

-   Flu A versus Flu B

-   Age group distributions

    -   broken down by Flu A versus Flu B

    -   broken down by CDC week

-   hospitalizations

    -   broken down by CDC week

    -   broken down by age

```{r}

#installing package for MEM method 
install.packages("mem", repos = "https://cloud.r-project.org")

#loading necessary packages 
library(tidyverse)
library(readr)
library(lubridate)
library(broom)
library(zoo)
library(ggplot2)
library(mem)

```

```{r}

#loading in PA ILINet data from 2014-2025
ILINet <- read_csv("C:/Users/hopeh/Downloads/Fall 2025/EPID 6000 Data Science for Biomedical Informatics/Final Project/FluViewPhase2Data_USE/ILINet.csv")

#keeping only relevant columns 
ILINet <- ILINet |>
  select("REGION TYPE", "REGION", "YEAR", "WEEK", "%UNWEIGHTED ILI", "ILITOTAL", "TOTAL PATIENTS" )

#renaming for clarity 
names(ILINet)[names(ILINet) == "%UNWEIGHTED ILI"] <- "percent.ili"

#checking if there are any missing values 
any(is.na(ILINet)) #no missing


```

According to the CDC, the "peak" months of influenza season are October through May \[7\]. We will use this timeline to create our influenza seasons. These dates correspond to CDC weeks 40-20. The ILINet data needs to be mutated to create influenza seasons that cross over two years.

```{r}

#Establishing Influenza Seasons for WHO

ILINet_clean <- ILINet |>
  mutate(
    #Creating a flu season variable
    flu.season = case_when(
      WEEK >= 40 ~ paste0(YEAR, "/", substr(YEAR + 1, 3, 4)), #Oct-Dec belong to season starting that year
      WEEK <= 20 ~ paste0(YEAR - 1, "/", substr(YEAR, 3, 4)), #Jan-May belong to to season that started previous year 
      TRUE ~ NA_character_                                        
    )
  ) |>
  
  #filter off-season weeks (weeks 21-39)
  filter(!is.na(flu.season)) |>
  
  #Check if each season has a week 53 (not typical, but happens in some calendar years)
  group_by(flu.season) |>
  mutate(
    has_week_53 = any(WEEK == 53)
  ) |>
  ungroup() |>
  
#Creating a variable for "week in flu season" - this will help with plotting later 
#It will be important to note the distinction between WEEK = CDC Week, and week.in.season = strictly for plotting
  mutate (
    week.in.season = case_when(
      WEEK >= 40 & WEEK <= 52 ~ WEEK - 39,    #Weeks 40-52 become 1-13
      WEEK == 53 ~ 14,                        #Week 53 becomes 14 (if it exists)
      WEEK <= 20 & has_week_53 ~ WEEK + 14,   #Weeks 1-29 become 15-34   OR
      WEEK <= 20 & !has_week_53 ~ WEEK + 13,  #Weeks 1-20 become 14-34 
      TRUE ~ NA_real_
    )
  ) |>
  #removing the help column
  select(-has_week_53) |>
  arrange(flu.season, week.in.season)

#View summary 
summary(ILINet_clean)


```

```{r}

#Establishing Influenza Seasons for MEM

ILINet_clean_mem <- ILINet |>
  mutate(
    #Creating a flu season variable
    flu.season = case_when(
      WEEK >= 40 ~ paste0(YEAR, "/", substr(YEAR + 1, 3, 4)), #Oct-Dec belong to season starting that year
      TRUE       ~ paste0(YEAR - 1, "/", substr(YEAR, 3, 4)), #All other weeks  
    )
  ) |>
  
  #Check if each season has a week 53 (not typical, but happens in some calendar years)
  group_by(flu.season) |>
  mutate(
    has_week_53 = any(WEEK == 53)
  ) |>
  ungroup() |>
  
#Creating a variable for "week in flu season" - this will help with plotting later 
#It will be important to note the distinction between WEEK = CDC Week, and week.in.season = strictly for plotting
  mutate (
    week.in.season = case_when(
      WEEK >= 40 & WEEK <= 52 ~ WEEK - 39,    #Weeks 40-52 become 1-13
      WEEK == 53 ~ 14,                        #Week 53 becomes 14 (if it exists)
      WEEK <= 20 & has_week_53 ~ WEEK + 14,   #Weeks 1-29 become 15-34   OR
      WEEK <= 20 & !has_week_53 ~ WEEK + 13,  #Weeks 1-20 become 14-34 
      WEEK >= 21 & WEEK <= 39 ~ NA_real_      #"Off-season" - no plotting index
    )
  ) |>
  #removing the help column
  select(-has_week_53) |>
  arrange(flu.season, week.in.season)

#View summary 
summary(ILINet_clean)

```

With the years available on ILINet, seasons 2014/15 through 2023/24 can be used to establish thresholds, and these will be validated on season 2024/25.

```{r}

#Creating a Validation Year
baseline.seasons <- ILINet_clean |>
  filter(flu.season %in% c("2014/15", "2015/16", "2016/17", "2017/18", 
                       "2018/19", "2019/20", "2020/21", "2021/22", 
                       "2022/23", "2023/24"))

validation.season <- ILINet_clean |>
  filter(flu.season == "2024/25") 

```

## Results {#sec-results}

Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

#### 1) Establishing thresholds using WHO method

The WHO method for defining thresholds was established in 2012 with the purpose of developing global standards for influenza surveillance \[3\]. It depends on historical data, and aligns previous years on the median week of peak activity. Then thresholds are assigned based on means and standard deviations of the aligned data. The paper by Tay et al. was used as a guide in applying this method to the ILINet data in Pennsylvania.

The WHO method requires identifying epidemic weeks to exclude them from baseline \[3\]. If we include epidemic weeks in our baseline, we would inflate the mean and standard deviation, creating thresholds that are too high. There are multiple ways to do this, with the simplest being a visual inspection of potential epidemic periods using a line graph. A more robust approach is the within-season percentile method. This will identify epidemic weeks as those are are unusually high relative to other weeks in the same season. Additionally, a moving average will be used to smooth data and reduce noise \[8\].

```{r}

### Step 1. Identifying Epidemic Periods 

epidemic.periods <- function(ILINet_clean, percentile = 0.90) {
  percentile.data <- ILINet_clean |>
    arrange(flu.season, week.in.season) |>
    group_by(flu.season) |>
    mutate(
      #calculating season-specific percentile thresholds
      season.threshold = quantile(percent.ili, probs = percentile, na.rm = TRUE),
      #Weeks above this threshold are considered epidemic weeks 
      above.thresh = percent.ili >= season.threshold,
      #Using moving average to smooth
      ma_3weeks = zoo::rollmean(percent.ili, k = 3, fill = NA, align = "center"), 
      #Identifying peak week in each season
      peak.week = percent.ili == max(percent.ili, na.rm = TRUE) )|>
        ungroup()
  
  #Create a summary for each season        
  epidemic.summary <- percentile.data |>
    group_by(flu.season) |>
    summarise(
      total.weeks = n(),
      epidemic.weeks = sum(above.thresh, na.rm = TRUE),
      percent.epidemic = round(100*epidemic.weeks / total.weeks, 1),
      peak.ili = max(percent.ili, na.rm = TRUE),
      peak.week = week.in.season[which.max(percent.ili)]
    )
  
  return(list(ILINet_clean = percentile.data, summary = epidemic.summary))
}
  
#Next, apply this within-season percentile method to our baseline seasons 
within.season.percent <- epidemic.periods(baseline.seasons, percentile = 0.90)

print(within.season.percent$summary)


############# MAKE THIS CLEANER ##############
##############################################
# Visualize identified epidemic periods
ggplot(within.season.percent$ILINet_clean, aes(x = week.in.season, y = percent.ili)) +
  geom_line(aes(group = flu.season), color = "gray60") +
  geom_point(aes(color = above.thresh), size = 2) +
  geom_point(data = filter(within.season.percent$ILINet_clean, peak.week), 
             color = "red", size = 4, shape = 17) + #Triangle for peak
  facet_wrap(~flu.season, scales = "free_y") +
  scale_color_manual(
    values = c("FALSE" = "blue", "TRUE" = "orange"),
    labels = c("Non-epidemic", "Epidemic")
  ) +
  theme_minimal() +
  labs(
    title = "Epidemic Period Identification (Within-Season 90th Percentile)",
    subtitle = "Red triangles indicate season peaks",
    x = "Week in Season",
    y = "Percent ILI (%)",
    color = "Classification"
  ) +
  theme(legend.position = "bottom")

```

Now, with "epidemic" weeks established, we can move on to exploring baseline (non-epidemic) activity.

```{r}

### Step 2: Calculating Baseline Activity 

baseline.activity <- within.season.percent$ILINet_clean |>
  filter(!above.thresh) |>   #exclude weeks above 90th percentile in each season
  summarise(
    baseline.mean = mean(percent.ili, na.rm = TRUE),  #calculate mean of baseline weeks
    baseline.sd = sd(percent.ili, na.rm = TRUE),      #calculate standard dev. of baseline weeks
    n.weeks = n()
  )


```

The WHO method defined three levels of flu thresholds: seasonal, average, and alert \[3\]. When the seasonal threshold is crossed into, this signals the start of flu season. We will define the seasonal threshold as mean + 2 SD (95th percentile), and the alert threshold as mean + 3 SD (99.7th percentile). Tay et al. explains that there is not a mandated approach to choosing these threshold definitions.

```{r}

### Step 3. Calculate WHO Thresholds

thresholds <- baseline.activity |>
  mutate(
    average.threshold = baseline.mean + (2 * baseline.sd),
    alert.threshold = baseline.mean + (3 * baseline.sd)
  )

print(thresholds)

```

Based on our historical flu season data (2014/15 through 2023/24), the seasonal threshold is 3.6% and the alert threshold is 4.5%. Note - these percentages represent the percent of outpatient medical visits associated with an ILI diagnosis in Pennsylvania.

```{r}

### Step 4. Visualize WHO Method Results 

#Applying thresholds to current flu season (2024/25)
if(nrow(validation.season) > 0) {
  current.flu.season <- ggplot() +
    #Current season (2024/25) data
    geom_line(
      data = validation.season,
      aes(x = week.in.season, y = percent.ili, group = flu.season),
      color = "black",
      alpha = 0.5
    ) +
    #Adding each threshold line individually
    geom_hline(yintercept = thresholds$baseline.mean, color = "blue", linetype = "dashed", size = 1) +
    geom_hline(yintercept = thresholds$average.threshold, color = "orange", linetype = "dashed", size = 1) +
    geom_hline(yintercept = thresholds$alert.threshold, color = "red", linetype = "dashed", size = 1)
}

print(current.flu.season)


###Still need to make this cleaner###
#For context - blue line is basically mean activity (seasonal threshold); orange line is average threshold, red line is alert threshold
```

-   code used to establish thresholds w/ explanation

-   graph of each year

#### 2) Establishing thresholds using MEM method

The MEM method is a similar yet slightly more sophisticated version of the WHO method. It uses automated algorithms to identify epidemic periods. This method was established by Vega and Lozano in 2013 and has been used previously to establish epidemic thresholds in European countries \[5\]. There is an R package available to download with a reference manual which was referred to throughout this project \[6\].

The MEM method has three main steps:

-   Determine epidemic timing.

    -   This means identifying the start of influenza season, how long it is, and when it ends for each historical season available.

-   Calculate baseline and epidemic thresholds.

    -   To do this, we use the previously established pre-epidemic and post-epidemic values.

-   Calculate intensity thresholds.

    -   This includes creating a threshold for each severity level (medium, high, very high).

A few key differences from the WHO method including using the Maximum Accumulated Percentage (MAP) curve instead of arbitrary percentiles, as well as providing more intensity levels.

```{r}

### Step 1: Prepping data in MEM Format 

ILINet_clean_mem <- ILINet_clean_mem |>
  filter(flu.season %in% c("2014/15", "2015/16", "2016/17", "2017/18", 
                           "2018/19", "2019/20", "2020/21", "2021/22", 
                           "2022/23", "2023/24"),
          WEEK != 53) |>        # remove week 53 
  select(flu.season, WEEK, percent.ili) |>
  arrange(flu.season, WEEK)


#To use MEM, we need our data in wide format (each season is a row)
mem_data <- ILINet_clean_mem |>
  pivot_wider(
    names_from = WEEK,
    values_from = percent.ili,
    names_prefix = "week_"
  ) |>
  #removing the season column, but keeping it for reference
  select(-flu.season) 



#add row names (season labels)
rownames(mem_data) <- ILINet_clean_mem |>
  distinct(flu.season) |>
  pull(flu.season)

#check for any NA values - if greater than 10% are missing in a season, it will need to be excluded to use MEM 
na_summary <- data.frame(
  season = rownames(mem_data),
  n_weeks = ncol(mem_data),
  n_missing = rowSums(is.na(mem_data)),
  pct_missing = round(100 * rowSums(is.na(mem_data)) / ncol(mem_data), 1)
)

na_summary  #there is no missing, so we can include all available seasons
```

Now, we will run the MEM model. The majority of the MEM analysis uses the "memmodel()" function. This identifies epidemic periods for each season, calculated the pre- and post- epidemic thresholds, and calculates the intensity thresholds.

```{r}

#Step 2. Running the MEM Model 

#Here we are running the MEM model with default parameters 
mem_results <- memmodel(
  i.data = mem_data,        #input data for the model     
  i.type.threshold = 1,     #calculating pre- and post- epidemic thresholds; option 1 is arithmetic mean 
  i.level.threshold = 0.95,
  i.tails.threshold = 1,  #uses 95th percentile of historical epidemic activity to set the threshold
  i.type.intensity = 6,      #calculating how severe an ongoing flu season is; option 6 = geometric mean and SDs
  i.level.intensity = c(0.50, 0.90, 0.975), 
  #low intensity = below 50thth percentile 
  #medium intensity = 40th-90th percentile 
  #high intensity = 90th-97.5th percentile
  #very high intensity = above 97.5th percentile 
  i.tails.intensity = 1,     #no tail adjustment
  i.type.curve = 2,          #normal curve
  i.level.curve = 0.95,      #significance level for the epidemic curve fit
  i.method = 1,              #estimation method for the epidemic threshold calculation = arithmetic/normal 
  i.param = 2.5,             #controls shape or spread of epidemic curve
  i.n.max = -1,              #consider all weeks  
  i.type.boot = "norm",      #bootstrap distribution used to estimate CIs for thresholds and intensities = normal
  i.iter.boot = 10000        #10,000 bootstrap iterations 
)

mem_results

### Step 2.5 Extract MEM Thresholds and Results 

mem_thresholds <- data.frame(
  threshold_type = c(
    "Epidemic Threshold",
    "Medium Intensity (50%)",
    "High Intensity (90%)",
    "Very High Intensity (97.5%)"
  ),
  value = c(
    mem_results$epidemic.thresholds[1],  # Epidemic Threshold
    mem_results$intensity.thresholds[1], # Medium
    mem_results$intensity.thresholds[2], # High
    mem_results$intensity.thresholds[3]  # Very High
  )
)

mem_thresholds


```

Vega et al. (2013) recommends using the geometric mean to calculate thresholds and intensities because epidemic peaks are inherently skewed. Here, a hybrid approach was used. The pre-epidemic thre

Since here we are working with state level data, our percent ILI values are smaller. The geometric mean is sensitive to values near zero - which is why in this project the arithmetic mean was used for the epidemic threshold and the geometric mean was used for intensities. 

When setting the intensity percentiles, we are telling the model to calculate the values based on those percentiles from our historical data. 

For example, for the medium intensity, the model is looking at the epidemic peaks from the 9 historical seasons provided and calculating the 50th percentile of those peaks. 

The mem model above provides an epidemic threshold, as well as three epidemic intensities. These percentages correspond to the percent of medical visits related to ILI diagnosis in any given week.

-   Epidemic Threshold: 2.06%

-   Medium Intensity: 2.45%

-   High Intensity: 4.17%

-   Very High Intensity: 5.50%

```{r}

###Step 3: Extracting Epidemic Timing 
pre_epidemic_threshold <- mem_thresholds$value[1]


epidemic_timing <- mem_data |>
  as.data.frame() |>
  mutate(season = rownames(mem_data)) |>
  pivot_longer(
    cols = starts_with("week_"),
    names_to = "cdc_week",
    values_to = "percent_ili",
    names_prefix = "week_"
  ) |>
  mutate(
    cdc_week = as.numeric(cdc_week),
    above_threshold = percent_ili >= pre_epidemic_threshold 
  ) |>
  group_by(season) |>
  summarise(
    #Epidemic start: first week above threshold 
    epidemic_start_cdc = ifelse(any(above_threshold), 
                               first(cdc_week[above_threshold]), 
                               NA),
    #Epidemic end: last consecutive week above threshold
     epidemic_end_cdc = ifelse(any(above_threshold), 
                             last(cdc_week[above_threshold]), 
                             NA),
    #Peak week and value
    peak_week_cdc = cdc_week[which.max(percent_ili)],
    peak_value = max(percent_ili, na.rm = TRUE),
    #Count weeks above threshold
    weeks_above_threshold = sum(above_threshold, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
  #Calculate duration (from first to last week above threshold)
  epidemic_duration = epidemic_end_cdc - epidemic_start_cdc + 1)


#Table for epidemic timing
epidemic_timing_table <- epidemic_timing |>
  select(
    Season = season,
    `Start (CDC Week)` = epidemic_start_cdc,
    `End (CDC Week)` = epidemic_end_cdc,
    `Duration (weeks)` = epidemic_duration,
    `Peak (CDC Week)` = peak_week_cdc,
    `Peak ILI (%)` = peak_value
  ) |>
  mutate(`Peak ILI (%)` = round(`Peak ILI (%)`, 2))

epidemic_timing_table


```


```{r}

###Step 4: Visualizing MEM Results

#Using built-in MEM plot
png("mem_model_plot.png", width = 12, height = 8, units = "in", res = 300)
plot(mem_results, main = "Pennsylvania Influenza Surveillance - MEM")


#Creating more custom visualizations
#reshaping data for plotting 
mem_data_long <- ILINet_clean_mem |>
  filter(flu.season %in% rownames(mem_data)) |>
  mutate(
    above_threshold = percent.ili >= mem_thresholds$value[1]
  ) |>
  # Join with epidemic timing to classify periods
  left_join(
    epidemic_timing |>
      select(season = season, epidemic_start_cdc, epidemic_end_cdc),
    by = c("flu.season" = "season")
  ) |>
  mutate(
    period = case_when(
      WEEK < epidemic_start_cdc ~ "Pre-epidemic",
      WEEK >= epidemic_start_cdc & WEEK <= epidemic_end_cdc ~ "Epidemic",
      WEEK > epidemic_end_cdc ~ "Post-epidemic",
      TRUE ~ "Unknown"
    ),
    period = factor(period, levels = c("Pre-epidemic", "Epidemic", "Post-epidemic"))
  )

##Plot 1: All Seasons With Thresholds##
plot_mem_all_seasons <- ggplot(mem_data_long, 
                               aes(x = WEEK, y = percent.ili)) +
  geom_line(aes(group = flu.season), color = "gray70", alpha = 0.5) +
  geom_point(aes(color = period), size = 1.5, alpha = 0.6) +
  
  # Add threshold lines
  geom_hline(yintercept = mem_thresholds$value[1], 
             color = "blue", linetype = "dashed", size = 1) +
  geom_hline(yintercept = mem_thresholds$value[3], 
             color = "yellow3", size = 0.8) +
  geom_hline(yintercept = mem_thresholds$value[4], 
             color = "orange", size = 0.8) +
  geom_hline(yintercept = mem_thresholds$value[5], 
             color = "red", size = 0.8) +
  
  # Add labels
  annotate("text", x = 30, 
           y = mem_thresholds$value[1],
           label = paste0("Pre-epidemic: ", round(mem_thresholds$value[1], 2), "%"),
           vjust = -0.5, color = "blue", size = 3) +
  annotate("text", x = 30,
           y = mem_thresholds$value[3],
           label = paste0("Medium: ", round(mem_thresholds$value[3], 2), "%"),
           vjust = -0.5, color = "yellow4", size = 3) +
  annotate("text", x = 30,
           y = mem_thresholds$value[4],
           label = paste0("High: ", round(mem_thresholds$value[4], 2), "%"),
           vjust = -0.5, color = "orange", size = 3) +
  annotate("text", x = 30,
           y = mem_thresholds$value[5],
           label = paste0("Very High: ", round(mem_thresholds$value[5], 2), "%"),
           vjust = -0.5, color = "red", size = 3) +
  
  scale_color_manual(
    values = c("Pre-epidemic" = "steelblue", 
               "Epidemic" = "coral", 
               "Post-epidemic" = "lightblue")
  ) +
  
  theme_minimal() +
  labs(
    title = "Pennsylvania Influenza Thresholds - MEM Method",
    subtitle = paste0("Based on ", length(unique(mem_data_long$flu.season)), " seasons"),
    x = "Week in Season (Week 1 = CDC Week 40)",
    y = "Percent ILI (%)",
    color = "Epidemic Period"
  ) +
  theme(legend.position = "bottom")

print(plot_mem_all_seasons)

```


```{r}

 ###Step 5: Applying MEM Thresholds to the Current Season

#Classify each week to a threshold
classification <- validation.season |>
  mutate(
    intensity_level = case_when(
      percent.ili < mem_thresholds$value[1] ~ "Below Threshold",
      percent.ili < mem_thresholds$value[3] ~ "Baseline",
      percent.ili < mem_thresholds$value[4] ~ "Medium",
      percent.ili < mem_thresholds$value[5] ~ "High",
      percent.ili >= mem_thresholds$value[5] ~ "Very High",
      TRUE ~ "Unknown"
    ),
    intensity_level = factor(
      intensity_level,
      levels = c("Below Threshold", "Baseline", "Medium", "High", "Very High")
    ),
    above_epidemic_threshold = percent.ili >= mem_thresholds$value[1]
  )

#Find epidemic alert
first_alert <- classification |>
  filter(above_epidemic_threshold) |>
  slice(1)

## Plot 2. Season 2024/25 With MEM Thresholds
plot_season_2024_2025 <- ggplot(classification, 
                           aes(x = week.in.season, y = percent.ili)) +
  # Shaded regions for intensity levels
  annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = 0, ymax = mem_thresholds$value[1],
           fill = "lightblue", alpha = 0.2) +
  annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = mem_thresholds$value[1], ymax = mem_thresholds$value[3],
           fill = "lightgreen", alpha = 0.2) +
  annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = mem_thresholds$value[3], ymax = mem_thresholds$value[4],
           fill = "yellow", alpha = 0.2) +
  annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = mem_thresholds$value[4], ymax = mem_thresholds$value[5],
           fill = "orange", alpha = 0.2) +
  annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = mem_thresholds$value[5], ymax = Inf,
           fill = "red", alpha = 0.2) +
  
  #Current season data
  geom_line(color = "black", size = 1.2) +
  geom_point(aes(color = intensity_level), size = 4) +
  
  #Add text labels with CDC week
  geom_text(aes(label = paste0("W", WEEK)), 
            vjust = -1, size = 2.5, color = "gray30") +
  
  #Threshold lines
  geom_hline(yintercept = mem_thresholds$value[1], 
             color = "blue", linetype = "solid", size = 1) +
  geom_hline(yintercept = mem_thresholds$value[3], 
             color = "yellow3", linetype = "dashed") +
  geom_hline(yintercept = mem_thresholds$value[4], 
             color = "orange", linetype = "dashed") +
  geom_hline(yintercept = mem_thresholds$value[5], 
             color = "red", linetype = "dashed") +
  scale_color_manual(
    values = c(
      "Below Threshold" = "steelblue",
      "Baseline" = "green4",
      "Medium" = "yellow3",
      "High" = "orange",
      "Very High" = "red"
    )
  ) +
  
  theme_minimal() +
  labs(
    title = "2024/25 Pennsylvania Influenza Season - MEM",
    subtitle = paste0("Epidemic threshold: ", round(mem_thresholds$value[1], 2), "% ILI"),
    x = "Week in Season (Week 1 = CDC Week 40)",
    y = "Percent ILI (%)",
    color = "Intensity Level",
    caption = "Labels show CDC week numbers"
  ) +
  theme(legend.position = "bottom")

print(plot_season_2024_2025)

```


3\) Some type of sensitivity analysis of two methods

-   don't necessarily have reference data

## Conclusion

This the conclusion. The @sec-results can be invoked here.

## References

1.  CDC. U.S. Influenza Surveillance: Purpose and Methods. FluView. March 3, 2025. Accessed October 29, 2025. https://www.cdc.gov/fluview/overview/index.html
2.  CDC. How CDC Classifies Flu Severity each Season in the United States. Influenza (Flu). November 26, 2024. Accessed October 28, 2025. https://www.cdc.gov/flu/php/surveillance/index.html
3.  Tay EL, Grant K, Kirk M, Mounts A, Kelly H. Exploring a Proposed WHO Method to Determine Thresholds for Seasonal Influenza Surveillance. Cook AR, ed. *PLoS ONE*. 2013;8(10):e77244. doi:10.1371/journal.pone.0077244
4.  *Global Epidemiological Surveillance Standards for Influenza*. World Health Organization; 2013.
5.  Vega T, Lozano JE, Meerhoff T, et al. Influenza surveillance in Europe: establishing epidemic thresholds by the moving epidemic method. *Influenza Other Respir Viruses*. 2013;7(4):546-558. doi:10.1111/j.1750-2659.2012.00422.x
6.  Lozano JE. mem: The Moving Epidemic Method. Published online November 26, 2010:2.19. doi:10.32614/CRAN.package.mem
7.  CDC. Flu Season. Influenza (Flu). September 26, 2025. Accessed November 3, 2025. https://www.cdc.gov/flu/about/season.html
8.  Steiner SH, Grant K, Coory M, Kelly HA. Detecting the start of an influenza outbreak using exponentially weighted moving average charts. *BMC Med Inform Decis Mak*. 2010;10(1):37. doi:10.1186/1472-6947-10-37
